---
title: Mapping Police Violence
author: Jessica Zhiyu Guo
date: Aug 7, 2023
description: The data is collected by Campaign Zero. An incident of "police violence" is defined by Campaign Zero as "any incident where a law enforcement officer (off-duty or on-duty) applies, on a civilian, lethal force resulting in the civilian being killed whether it is considered 'justified' or 'unjustified' by the U.S. Criminal Legal System."
categories:
  - GLMs
  - linear regression
  - logistic regression
data:
  year: 2023
  files: 
  - mapping-police-violence.csv
  - cc-est2022-all.csv
---

## Motivation
Race and policing are some of the most inescapable topics in today's America. Mapping Police Violence dataset aims to provide a fuller landscape of police violence, expanding the definition to include various causes of death and beyond the scope of previous governmental and other databases of a similar kind (eg. Fatal Force/The Washington Post). It is the most comprehensive dataset detailing all kinds of police violence incidents and helps us understand the scale of police violence and the differences across race/ethnicity and different geographical locations. We can also combine this with the 2020 Census data on the county level to understand the prevalence of police violence and racial/ethnic differences in prevalence in different counties. The methodology is last updated on Oct 3, 2022, and the dataset is downloaded on Aug 7, 2023.


## Data
Each row represents one victim in an incident of police violence

Mapping Police Violence include detailed descriptions of each incident, including date and victim's age, race, gender, and an image. It also includes more details about the incident including the location to the street address, the cause of death (can be multiple), circumstances, agency responsible, initial reasons, officer names and history, and so on. 

Not all variables are available for all incidents, for older incidents, some data around prosecution (and its result) for the officer could be available.


## Data preview

```{r, echo=FALSE, results="asis"}
source("../preview_dataset.R")
preview_datasets()
```

## Variable Descriptions
The [dataset codebook from Mapping Police Violence](https://mappingpoliceviolence.org/files/MappingPoliceViolence_Methodology.pdf) and the [U.S. Census Bureau documentation](https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2022/cc-est2022-alldata.pdf) contain the full list of variables. This is an extract of a few useful variables from both tables. Note that many variables can have multiple values, separated by commas in the Mapping Police Violence dataset.


#### Mapping Police Violence
| Variable | Description |
|----|-------------|
| name | 	Name of victim killed by law enforcement |

| age | Age of victim at time of death |
| gender  |	Victim's indicated gender by news/official reports |

| race  |	Race of victim according to news/official reports |
| date |	Date of lethal action on victim |
| city |	City where incident took place |

| state  |	State where incident took place |
| county  |	County where incident took place |

| agency_responsible|	Agency of law enforcement officer who took lethal action and responsible for killing |
| cause_of_death |	Highest level of force used by law enforcement officer: Gun/Taser/Taser, Gunshot/Unknown/Vehicle/Vehicle, Gunshot/Other Physical Force|
| signs_of_mental_illness|	Indicates whether official/news reports mentioned any signs of mental illness or mental health crises: Yes/No/Unknown/Drug or Alcohol Use|
| allegedly_armed |	Indicates whether victim was armed according to news/official sources: Allegedly Armed/Unarmed/Unclear/Vehicle |
| off_duty_killing |	Indicates whether officer was working in an official capacity at the time lethal action was taken |
| encounter_type |	Initial reported reason(s) for police to be on the scene prior to using deadly force |
| officer_names  |	Name of officer responsible for highest level of force|
| officer_races  |	Race of officer responsible for highest level of force |
| officer_known_past_shootings |	Indicates if officer has had a previous incident of police violence in the database |

#### Census (CC-EST2022-ALLDATA-[ST-FIPS]: Annual County Resident Population Estimates by Age, Sex, Race, and Hispanic Origin: April 1, 2020 to July 1, 2022)
| Variable | Description |
|----|-------------|
| STNAME | 	State Name  |
| CTYNAME| County Name  |
| NHWA_MALE/NHWA_FEMALE  | Not Hispanic, White alone male/female population |
| NHBA_MALE/NHBA_FEMALE  |	Not Hispanic, Black or African American alone male/female population  |
| NHIA_MALE/NHIA_FEMALE |	Not Hispanic, American Indian and Alaska Native alone male/female population|
| NHAA_MALE/NHAA_FEMALE| Not Hispanic, Asian alone male/female population  |
| NHNA_MALE/NHNA_FEMALE  |Not Hispanic, Native Hawaiian and Other Pacific Islander alone male/female population |
| NHTOM_MALE/NHTOM_FEMALE  | Not Hispanic, Two or More Races male/female population |
| H_MALE/H_FEMALE|	Hispanic male/female population|

## Questions
1. Do some EDA. What's the age and gender distribution of victims? What are the most common encounter types in the dataset? What are the most common causes of death? Explore these questions, and others that seem interesting in the data, with charts and graphs.
2. One main question we would like to explore is that whether the demographic composition of the locality affects the occurrences of police violence. We have incorporated demographic data from the census on the county level with 7 groups (non-Hispanic White, non-Hispanic Black, non-Hispanic Asian, non-Hispanic American Indian/Alaskan Native, non-Hispanic Native Hawaiian or other Pacific Islanders, non-Hispanic two or more races, and Hispanic) with the code below. Build a model predicting the count of police violence in each county in the dataset based on their racial/ethnic minority percentage and other factors you think might be relevant, what does the result suggest?
3. Another argument people often make is that officers responsible for police violence against minorities often bear no consequences for their actions. Based on this data, build a logistic model to predict the likelihood of the police officer(s) responsible being charged with all factors you think might be relevant, what does the result suggest?
4. Some people suggest that sign(s) of mental health issues and/or drug use might also affect how police respond to a person, say we categorize any mentioning of a gunshot in the cause of death as an indicator for a higher level of force used, does our data seem to support this claim? 
5. Similarly, does the perception of the individual being armed appear to influence the extent of force employed by the police?

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(dplyr)
mpv <- read.csv("Mapping Police Violence.csv", sep = ",", header = T)
colnames(mpv)[1] <- c("name")
cc <- read.csv("cc-est2022-all.csv", sep = ",", header = T)
cc <- cc %>% filter(YEAR == "1") %>% # selecting base year (2020)
  select(STNAME, CTYNAME, NHWA_MALE,NHWA_FEMALE, #non hisp, white alone
                                            NHBA_MALE,NHBA_FEMALE, #nh, black alone
                                            NHAA_MALE, NHAA_FEMALE, #nh, asian alone
                                            NHIA_MALE,NHIA_FEMALE, #nh, native alone
                                            NHNA_MALE,NHNA_FEMALE, #nh, pacific is alone
                                            NHTOM_MALE,NHTOM_FEMALE, #nh, multiple races
                                            H_MALE, H_FEMALE) %>% #hispanic
  mutate(WA = NHWA_MALE+NHWA_FEMALE, BA = NHBA_MALE+NHBA_FEMALE,
         AA = NHAA_MALE+NHAA_FEMALE, IA = NHIA_MALE+NHIA_FEMALE,
         NP = NHNA_MALE+NHNA_FEMALE, H = H_MALE+H_FEMALE) %>% 
  select(WA, BA, AA, IA, NP, H, STNAME, CTYNAME)


cc$CTYNAME = gsub(" Parish", "", cc$CTYNAME) #removing Parish from names
cc$CTYNAME = gsub(" County", "", cc$CTYNAME) #removing County 
cc$CTYNAME = gsub(" Borough", "", cc$CTYNAME) #removing Borough (AK)
cc$CTYNAME = gsub(" Census Area", "", cc$CTYNAME) #removing Census Area (AK)
cc$CTYNAME = gsub("-", " ", cc$CTYNAME) #changing - to space

mpv$county = gsub(" Parish", "", mpv$county) #removing Parish from names
mpv$county = gsub(" County", "", mpv$county) #removing County
mpv$county = gsub(" Census Area","", mpv$county) #removing Census Area (AK)
mpv$county = gsub(" Borough", "", mpv$county) #removing Borough (AK) 
mpv$county = gsub("-", " ", mpv$county) #changing - to space


mpv$state = state.name[match(mpv$state, state.abb)]

df = left_join(mpv, cc, by = c("state" = "STNAME", "county" = "CTYNAME"), 
               multiple = "any")

```

## References
- https://mappingpoliceviolence.org/, data retrieved on Aug 7, 2023.
- https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2022/cc-est2022-alldata.pdf
